---
title: "MS WORD Example"
output:
  word_document:
    pandoc_args:
    - --csl
    - stroke.csl
    - --bibliography
    - test.bib
    fig_caption: yes
  pdf_document: default
---

<!---       "--csl=elsevier-with-titles.csl" -->
<!---       "--csl=elsevier-harvard.csl" -->

```{r label=opts, results='hide', echo=FALSE, message = FALSE, warning=FALSE}
library(knitr)
opts_chunk$set(echo=TRUE, prompt=FALSE, message=FALSE, warning=FALSE, comment="", results='hide')

```

```{r, results = 'asis'}
library(knitcitations)
cleanbib()
cite_options(citation_format= "pandoc")

library(pander)
pander(summary(cars))
```


```{r, results = 'asis'}
ctl <- c(4.17,5.58,5.18,6.11,4.50,4.61,5.17,4.53,5.33,5.14)
trt <- c(4.81,4.17,4.41,3.59,5.87,3.83,6.03,4.89,4.32,4.69)
group <- gl(2, 10, 20, labels = c("Ctl","Trt"))
weight <- c(ctl, trt)
lm.D9 <- lm(weight ~ group)
pander(lm.D9)
```

And you can embed things coefficients like we always could using the `r ` syntax.  Like the coefficient for treatment was `r lm.D9$coefficients["groupTrt"]` and the CI was ....

```{r makefigcap}
figcap.cars = "This is my caption of things"
```

You can also embed plots, for example:

```{r, echo=FALSE, fig.cap = figcap.cars}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

$$
\alpha + \beta = x
$$

$$
  \begin{eqnarray}
  \dot{x} + \bar{x} &=& 2\\
  \hat{y} &=& 400
  \end{eqnarray}
$$

I think the `knitr` package `r citep(citation("knitr"))` is really awesome!  
I cannot believe this shit.  I can't believe `r citet(citation("knitcitations"))` really did it.

```{r bib, comment = "", results='markup'}
write.bibtex(file = "test.bib")
# bibliography()
```

<!---https://github.com/citation-style-language-->