```{r label=opts, results='hide', echo=FALSE, message = FALSE, warning=FALSE}
library(knitr)
opts_chunk$set(echo=TRUE, prompt=FALSE, message=FALSE, warning=FALSE, comment="", results='hide')
rm(list=ls())
```

# Manipulate Package
The [`manipulate`](http://cran.r-project.org/web/packages/manipulate/index.html) from RStudio allows you to create simple Tcl/Tk operators for interactive visualization.  I will use it for a simple slider to view different slices of an image.

```{r}
library(manipulate)
```

# fslr package
I'm calling the `fslr` package because I know that if you have it installed, you have `fsl` and have a 1mm T1 template from MNI in a specific location.  It also loads the `oro.nifti` package so that `readNIfTI` is accessible after loading `fslr`.  

Here I will read in the template image:
```{r}
library(fslr)
options(fsl.path='/usr/local/fsl')
template = file.path(fsldir(), "data/standard", 
                     "MNI152_T1_1mm_brain.nii.gz")
img = readNIfTI("~/Desktop/scratch/100-318_20070723_0957_CT_3_CT_Head-.nii.gz")
```

# The iplot function

The `iplot` function defined below takes in a `nifti` object, the specific plane to be plotted and additional options to be passed to `oro.nifti::image`.  

```{r}
iplot = function(img, plane = c("axial", 
                                "coronal", "sagittal"), ...){
  plane = match.arg(plane, c("axial", 
                             "coronal", "sagittal"))
  ns=  switch(plane,
              "axial"=dim(img)[3],
              "coronal"=dim(img)[2],
              "sagittal"=dim(img)[1])
  manipulate({
    image(img, z = z, plot.type= "single", plane = plane, ...)
    pos <- manipulatorMouseClick()
    if (!is.null(pos)) {
      print(pos)
    }
  },
  z = slider(1, ns, step=1, initial = ceiling(ns/2))
  )
}
```

The result will be a plotted image and the 

# Example plots

```{r}
iplot(img)
iplot(img, plane = "coronal")
iplot(img, plane = "sagittal")
```