```{r label=opts, results='hide', echo=FALSE, message = FALSE, warning=FALSE}
library(knitr)
opts_chunk$set(echo=FALSE, prompt=FALSE, message=FALSE, warning=FALSE, comment="", results='hide')
```

TL;DR - Neuroimaging-specific post.  Some files can't be converted by programs if they are encoded one way - present Matlab program to fix that.

## JPEG2000 DICOM data
Recently, I had some DICOM data that is stored in the JPEG2000 format from [OsiriX](http://www.osirix-viewer.com/datasets/).  I wanted to convert these to a NIfTI format using [dcm2nii](http://www.mccauslandcenter.sc.edu/mricro/mricron/dcm2nii.html) but it was not a fan of the compression, specifically the transfer syntax (which tells you how the data is encoded).

It spit out the error:
```
Unsupported Transfer Syntax 1.2.840.10008.1.2.4.91 Solution: use MRIcro
```

## Using Matlab's Image Processing Toolbox
For most of my endeavors, I try to use `R` for everything, but that's not always possible.  For example the `oro.dicom` package is great for working with DICOM data, but the JPEG2000 compression format is not supported.  Therefore, I went into Matlab, which has the [image processing toolbox](http://www.mathworks.com/products/image/).  You will need an updated version, [as described here, Matlab 2009b version will not work](http://www.mathworks.com/matlabcentral/answers/93596-why-is-dicomread-unable-to-read-my-jpeg-2000-encoded-image-using-the-image-processing-toolbox-6-4-r).  

## The Code for Conversion
Overall, the code takes a directory, lists all the files, removing directories.  .  The `for loop` goes through each DICOM file, reads in the header information with `dicominfo`, reads in the data matrix with `dicomread`.  The transfer syntax is changed using `newsyntax`, creates a new filename (within a specified output directory `outdir`) and writes the DICOM file using `dicomwrite`.  The `createmode` can either be `Create` or `Copy`.  Using `Create` is the default option and has error and missing checking.  The `Copy` mode bypasses many of these checks, but this may be required for certain DICOM files to be written.  


```{r, eval=FALSE, echo=TRUE}
rundir = pwd;
updir = fileparts(rundir);
outdir = fullfile(updir, 'Converted');
addon = '';
newsyntax = '1.2.840.10008.1.2';
createmode = 'Copy'; % could be 'Create';
x = dir(rundir);
nodir = ~[x.isdir]';
x = x(nodir);
x = {x.name}';
ifile = 1;
for ifile = 1:length(x)
    name = x{ifile};
    stub = regexprep(name, '(.*)[.].*', '$1');
    stub = [stub addon '.dcm']; %#ok<AGROW>
    name = fullfile(rundir, name);
    stub = fullfile(outdir, stub);
    dinfo = dicominfo(name);
    dinfo.TransferSyntaxUID = newsyntax;
    X = dicomread(name);
    dicomwrite(X, stub, dinfo, 'CreateMode', createmode);
end
```

After this conversion, `dcm2nii` worked as usual for converting. 

### Other Options
Other options exist such as the [DICOM JPEG 2000 Module](http://dicom.offis.de/dcmjp2k.php.en), but this must be licensed and Matlab has implemented this in `imread`.  Just figured if you ran into this problem, this may help.  