```{r label=opts, results='hide', echo=FALSE, message = FALSE, warning=FALSE}
library(knitr)
opts_chunk$set(echo=TRUE, prompt=FALSE, message=FALSE, warning=FALSE, comment="", results='hide')
opts_knit$set(error=TRUE)
```

In my [last post](http://hopstat.wordpress.com/2014/10/16/ggplot2-is-not-always-the_answer-its-not-supposed-to-be-7/), I discussed how `ggplot2` is not always the answer to the question "How should I plot this" and that base graphics were still very useful.

# Why Do I use ggplot2 then?
The overall question still remains: why (do I) use `ggplot2`?

## ggplot2 vs lattice

For one, `ggplot2` replaced the `lattice` package for many plot types for me.  The `lattice` package is a great system, but if you are plotting multivariate data, I believe you should choose `lattice` or `ggplot2`. I chose `ggplot2` for the syntax, added capabilities, and the philosophy behind it.  The fact [Hadley Wickham](http://had.co.nz/) is the developer never hurts either.  

## Having multiple versions of the same plot, with slight changes
Many times I want to do the same plot over and over, but vary one aspect of it, such as color of the points by a grouping variable, and then switch the color to another grouping variable.  Let me give a toy example, where we have an `x` and a `y` with two grouping variables: `group1` and `group2`. 

```{r data_gen}
library(ggplot2)
set.seed(20141016)
data = data.frame(x = rnorm(1000, mean=6))
data$group1 = rbinom(n = 1000, size =1 , prob =0.5)
data$y = data$x * 5 + rnorm(1000)
data$group2 = runif(1000) > 0.2
```

We can construct the `ggplot2` object as follows:
```{r g_create}
g = ggplot(data, aes(x = x, y=y)) + geom_point()
```

The `ggplot` command takes the `data.frame` you want to use and use the `aes` to specify which aesthetics we want to specify, here we specify the `x` and `y`.  Some aesthetics are optional depending on the plot, some are not.  I think it's safe to say you always need an `x`.  I then "add" (using `+`) to this object a "layer": I want a geometric "thing", and that thing is a set of points, hence I use `geom_points`.  I'm doing a scatterplot.

If you just call the object `g`, `print`  is called by default, which plots the object and we see our scatterplot.
```{r print_g, cache=TRUE}
g
```

I can color by a grouping variable and we can add that aesthetic:
```{r color_group, cache=TRUE}
g + aes(colour = group1)
g + aes(colour = factor(group1))
g + aes(colour = group2)
```

Note, `g` is the original plot, and I can add `aes` to this plot, which is the same as if I did `ggplot2(data, aes(...))` in the original call that generated `g`.  **NOTE** if the `aes` you are adding was not a column of the `data.frame` when you created the plot, you will get an error.  For example, let's add a new column to the data and then add it to `g`:

```{r g2_create}
data$newcol = rbinom(n = nrow(data), size=2, prob = 0.5)
g + aes(colour=factor(newcol))
```

This fails because the way the `ggplot2` object was created.  If we had added this column to the data, created the plot, then added the `newcol` as an `aes`, the command would work fine.

```{r g2_create2}
g2 = ggplot(data, aes(x = x, y=y)) + geom_point()
g2 + aes(colour=factor(newcol))
```



We see in the first plot with `colour = group1`, `ggplot2` sees a numeric variable `group1`, so tries a continuous mapping scheme for the color.  The default is to do a range of blue colors denoting intensity.   If we want to force it to a discrete mapping, we can turn it into a factor `colour = factor(group1)`.  We see the colors are very different and are not a continuum of blue, but colors that separate groups better.  
The third plot illustrates that when `ggplot2` takes logical vectors for mappings, it factors them, and maps the group to a discrete color.  

### Slight Changes with additions
In practice, I do this iterative process many times and the addition of elements to a common template plot is very helpful for speed and reproducing the same plot with minor tweaks.  

In addition to doing similar plots with slight grouping changes I also add different lines/fits on top of that.  In the previous example, we colored by points by different grouping variables.  In other examples, I tend to change little things, e.g. a different smoother, a different subset of points, constraining the values to a certain range, etc.  I believe in this way, `ggplot2` allows us to create plots in a more structured way, without copying and pasting the entire command or creating a user-written wrapper function as you would in base.  This should increase reproducibility by decreasing copy-and-paste errors.  I tend to have many copy-paste errors, so I want to limit them as much as possible.

## ggplot reduces number of commands I have to do
One example plot I make frequently is a scatterplot with a smoother to estimate the shape of bivariate data.  

In base, I usually have to run at least 3 commands to do this, e.g. `loess`, `plot`, and `lines`.  In `ggplot2`, `geom_smooth()` takes care of this for you.   Moreover, it does the smoothing by each different aesthetics (aka smoothing per group), which is usually what I want do as well (and takes more than 3 lines in base, usually a `for` loop or `apply` statement). 

```{r smooth}
g2 + geom_smooth()
```
By default in `geom_smooth`, it includes the standard error of the estimated relationship, but I usually only look at the estimate for a rough sketch of the relationship.  Moreover, if the data are correlated (such as in longitudinal data), the standard errors given by default methods are usually are not accurate anyway.  

```{r smooth_no_se}
g2 + geom_smooth(se = FALSE)
```

Therefore, on top of the lack of copying and pasting, you can reduce the number of lines of code.  Some say "but I can make a function that does that" -- yes you can.  Or you can use the commands already there in `ggplot2`. 

### Faceting
The other reason I frequently use `ggplot2` is for faceting.  You can do the same graph, conditioned on levels of a variable, which I frequently used.  Many times you want to do a graph, subset by another variable, such as treatment/control, male/female, cancer/control, etc.  
```{r facet_group, cache=TRUE}
g + facet_wrap(~ group1)
g + facet_wrap(~ group2)
g + facet_wrap(group2 ~ group1)
g + facet_wrap( ~ group2 + group1)
g + facet_grid(group2 ~ group1)
```


## Spaghetti plot with Overall smoother
I also frequently have longitudinal data and make spaghetti plot for a per-person trajectory over time.  
For this example I took code from [StackOverflow](http://stats.stackexchange.com/questions/76999/simulating-longitudinal-lognormal-data-in-r) to create some longitudinal data.


```{r sim_long}
library(MASS)
library(nlme)

### set number of individuals
n <- 200

### average intercept and slope
beta0 <- 1.0
beta1 <- 6.0

### true autocorrelation
ar.val <- .4

### true error SD, intercept SD, slope SD, and intercept-slope cor
sigma <- 1.5
tau0  <- 2.5
tau1  <- 2.0
tau01 <- 0.3

### maximum number of possible observations
m <- 10

### simulate number of observations for each individual
p <- round(runif(n,4,m))

### simulate observation moments (assume everybody has 1st obs)
obs <- unlist(sapply(p, function(x) c(1, sort(sample(2:m, x-1, replace=FALSE)))))

### set up data frame
dat <- data.frame(id=rep(1:n, times=p), obs=obs)

### simulate (correlated) random effects for intercepts and slopes
mu  <- c(0,0)
S   <- matrix(c(1, tau01, tau01, 1), nrow=2)
tau <- c(tau0, tau1)
S   <- diag(tau) %*% S %*% diag(tau)
U   <- mvrnorm(n, mu=mu, Sigma=S)

### simulate AR(1) errors and then the actual outcomes
dat$eij <- unlist(sapply(p, function(x) arima.sim(model=list(ar=ar.val), n=x) * sqrt(1-ar.val^2) * sigma))
dat$yij <- (beta0 + rep(U[,1], times=p)) + (beta1 + rep(U[,2], times=p)) * log(dat$obs) + dat$eij
```

I will first add an alpha level to the plotting lines for the next plot (remember this must be done before the original plot is created).  `tspag` will be the template plot, and I will create a spaghetti plot (`spag`) where each colour represents an `id`:
```{r spag, cache=TRUE}
library(plyr)
dat = ddply(dat, .(id), function(x){
  x$alpha = ifelse(runif(n = 1) > 0.9, 1, 0.1)
  x$grouper = factor(rbinom(n=1, size =3 ,prob=0.5), levels=0:3)
  x
})
tspag = ggplot(dat, aes(x=obs, y=yij)) + 
  geom_line() + guides(colour=FALSE) + xlab("Observation Time Point") +
  ylab("Y")
spag = tspag + aes(colour = factor(id))
spag
```

Many other times I want to group by `id` but plot just a few lines (let's say 10% of them) dark and the other light, and not colour them:

```{r, cache=TRUE}
bwspag = tspag + aes(alpha=alpha, group=factor(id)) + guides(alpha=FALSE)
bwspag
```

Overall, these 2 plots are useful when you have longitudinal data and don't want to loop over ids or use `lattice`.  The great addition is that all the faceting and such above can be used in conjunction with these plots to get spaghetti plots by subgroup.

```{r spag_facet, cache=TRUE}
spag + facet_wrap(~ grouper)
```

### Spaghetti plot with overall smoother
If you want a smoother for the overall group in addition to the spaghetti plot, you can just add `geom_smooth`:
```{r spag_smooth, cache=TRUE}
sspag = spag + geom_smooth(se=FALSE, colour="black", size=2)
sspag
sspag + facet_wrap(~ grouper)
bwspag + facet_wrap(~ grouper)
```

Note that the `group` aesthetic and `colour` aesthetic do not perform the same way for some operations.  For example, let's try to smooth `bwswag`:

```{r spag_smooth_bad, cache=TRUE}
bwspag + facet_wrap(~ grouper) + geom_smooth(se=FALSE, colour="red")
```

We see that it smooths each id, which is not what we want.  We can achieve the desired result by setting the `group` aesthetic:
```{r spag_smooth_corr, cache=TRUE}
bwspag + facet_wrap(~ grouper) + 
  geom_smooth(aes(group=1), se=FALSE, colour="red", size =2)
```

I hope that this demonstrates some of the simple yet powerful commands `ggplot2` allows users to execute.  I agree that some behavior may not seem straightforward at first glance, but becomes more understandable as one uses `ggplot2` more.  

## Another Case this is Useful - Save Plot Twice
Another (non-plotting) example I want to show is how saving `ggplot2` objects can make saving duplicate plots much easier.  In many cases for making plots to show to others, I open up a PDF device using `pdf`, make a series of plots, and then close the PDF. There may be 1-2 plots that are the real punchline, and I want to make a high-res PNG of them separately.  Let me be clear, I want both -- the PDF and PNG.  For example:

```{r pdf_and_pngs}
pdf(tempfile())
print({g1 = g + aes(colour = group1)})
print({g1fac = g + aes(colour = factor(group1))})
print({g2 = g + aes(colour = group2)})
dev.off()
png(tempfile(), res = 300, height =7, width= 7, units = "in")
print(g2)
dev.off()
```

I am printing the objects, while assigning them.  (I have to use the `{` brackets because I use `=` for assignment and `print` would evaluate that as arguments without `{}`).  As `g2` is already saved as an object, I can close the PDF, open a `png` and then print that again.

No copying and pasting was needed for remaking the plot, nor some weird turning off and on devices.  


## Conclusion
I (and think you should) use `ggplot2` for a lot of reasons, especially the grammar and philosophy.  The plots I have used here are some powerful representations of data that are simple to execute.  I believe using this system reflects and helps the true iterative process of making figures.  The syntax may not seem intuitive to a long-time `R` user, but I believe the startup cost is worth the effort.  Let me know if you'd like to see any other plots that you commonly use. 