---
output: html_document
---
```{r label=opts, results='hide', echo=FALSE, message = FALSE, warning=FALSE}
library(knitr)
opts_chunk$set(echo=TRUE, prompt=FALSE, message=TRUE, warning=TRUE, comment="")
setwd("~/Dropbox/Public/WordPress_Hopstat/Working_with_NIfTI_images_in_R/")
library(fslr)
```

The `oro.nifti` package is awesome for Neu`R`oimaging (couldn't help myself).  It has functions to read/write images, introduces the S4 `nifti` class, and has useful plotting functions.  There are some limitations and some gotchas that are important to discuss if you are working with these objects in `R`.  

## Dataset Creation

We'll read in some data (a template, example taken from `readNIfTI` in `oro.nifti`.  We have to download it as most packages aren't allowed to include data (Bioconductor has mostly remedied this for bioinformatics).

```{r makedata, cache=TRUE}
library(oro.nifti)
URL <- "http://imaging.mrc-cbu.cam.ac.uk/downloads/Colin/colin_1mm.tgz"
urlfile <- "colin_1mm.tgz"
if (!file.exists(urlfile)){
  download.file(URL, dest=urlfile, quiet=TRUE)
  untar(urlfile)
}
img <- readNIfTI("colin_1mm")
img = cal_img(img)
```

The `img` object read in is of class `nifti`, has a series of slots (remember S4 object, not S3), and the `print` method produces information about the NIfTI header.

```{r printimg}
class(img)
print(img)
slotNames(img)
hist(img)
```

### Array operation

Now, we can do simple operations on this object like any other array.  The problem is that certain operations do not (some rightfully so) return a `nifti` object.

For example, let's get an indicator if the value is greater than 400000.  (This is an MRI and has arbitrary units).


```{r}
mask = img > 400000
mask
class(mask)
```

We see a `nifti` object is returned, with all the header information contained.  Great!  Let's get an indicator of values greater than 400000 but less than 600000.  

```{r}
mask_2 = img > 400000 & img < 600000
class(mask_2)
```

We see that this operation returned an `array`, not a `nifti` object.  That's kind of unexpected, but probably conservative.  For example, if this had been 2 different `nifti` objects and you were performing the subsetting, which object's information should you use?

We see the same behavior when multiplying (or subtracting, etc) two `nifti` objects.

```{r}
img_masked = img * mask
class(img_masked)
```

We know the object is an `array` but we may want a `nifti` object. 

### `fslr` Helper functions
The `niftiarr` function in `fslr` inputs a `nifti` object and an `array`, and returns a `nifti` object with the array in the `@.Data` slot, copying over the image header information from the input `nifti` object.  

```{r niftiarr}
### need the development version
# library(devtools)
# devtools::install_github("muschellij2/fslr")
library(fslr)
img_masked = niftiarr(img, img * mask)
class(img_masked)
```

### Other ways to skin the cat

Another way of masking the image is to subset the values of the image that are not in the mask and setting those values to $0$ (or some other value).
```{r masked_2}
img_masked_2 = img
img_masked_2[mask == 0] = 0
class(img_masked_2)
```

We see that this correctly returns an object of class `nifti`.  One potential issue with this direction is that `cal_min` and `cal_max` slots on the resulting `nifti` object not the range of the image.  This is problematic because `orthographic` and `image.nifti` both use this slot to determine grayscale ranges for plotting and `writeNIfTI` errors if these values don't match.  The `cal_img` function in `fslr` is a simple helper function that will set these values to the respective values of the range of the data.

```{r cal_img}
range(img_masked_2)
c(cal.min(img_masked_2), cal.max(img_masked_2))
img_masked_2 = cal_img(img_masked_2)
range(img_masked_2)
```

Note, sometimes `readNIfTI` sets the `cal_min` and `cal_max` slots both to $0$, which may be a result of how the `nifti` was created.

#### Equivalent Operations
We see that after these operations done 2 different ways, the resulting `nifti` objects are equivalent.
```{r allequal}
all.equal(img_masked, img_masked_2)
```

## Data Typing and Writing `nifti` objects
I've worked with `oro.nifti` for some time, and I wanted to also discuss writing `nifti` objects.  Many operations preserve the data `type` of the array in the `nifti` object, such as if the data were integers, an array of integers returned.  Other times, the data type changes, such as if you smoothed an image of integers, the resulting `array` would have values that are not integers, but probably continuous values.  When plotting the data or operating with the `R` object, this is not generally an issue because any extraction of the information will take the `nifti` object's `@.Data` slot and return that array.

This presents a problem when you write `nifti` objects using the `writeNIfTI` function.  

## Slots `datatype` and `bitpix`
In a `nifti` object, the [`datatype`](http://nifti.nimh.nih.gov/nifti-1/documentation/nifti1fields/nifti1fields_pages/datatype.html) and [`bitpix`](http://nifti.nimh.nih.gov/nifti-1/documentation/nifti1fields/nifti1fields_pages/bitpix.html) fields specify what type the data is and the level of precision stored in the NIfTI file.  For example, if you have binary data, the 
`datatype` would be a value of 2 for `UINT8` (unsigned integer) and the `bitpix` would be 8 for 8 bits per voxel.  These must correspond to each other (i.e. no mismatching).  Also, storing data in a type that is what I call "over-storing" (e.g storing binary data in a `FLOAT32` format) may make files larger than necessary.  Under-storing is much more problematic, however.

### Why you should care about under-storing
If you read in data as integers, did some operations that returned continuous (decimal) objects, and then write the `nifti` object using the original header information would truncate the data in ways you wouldn't want.  In `R`, the object would look fine, but when you wrote it to disk, it would be truncated.

```{r}
img@bitpix
img@datatype
```

Let's say we scaled the data by its maximum:

```{r pct}
pct = img / max(img)
pct = cal_img(pct)
pct@bitpix
pct@datatype
hist(pct)
### we will explain this in the next section
pct = drop_img_dim(pct)
pct = zero_trans(pct)
pct = onefile(pct)

```

If we write this to disk, and read the result back in, we see:
```{r readwrite, fig.height=7, fig.width=2.3, dev='png', res = 300}
par(mfrow=c(3,1))
hist(pct, main="Correct Data")

### Writing out pct without any datatype change - still UINT8
tfile = tempfile()
writeNIfTI(pct, filename = tfile, verbose=TRUE)
pct2 = readNIfTI(paste0(tfile, ".nii.gz"))
hist(pct2, main="Writing out with no change of datatype/bitpix")

### Writing out pct with datatype change - to FLOAT32
pct = datatyper(pct, type_string = "FLOAT32")
writeNIfTI(pct, filename = tfile)
pct3 = readNIfTI(paste0(tfile, ".nii.gz"))
hist(pct3, main="Writing out with changing of datatype/bitpix")
```

As you can see, you **need** to change the bitpix and datatype when writing out NIfTI files if you are changing the data type in the array of data in `R`.  We had changed integers to percentages/proportions, and we changed the output to `FLOAT32` (for floating point data).  Take a look at `convert.datatype()` and `convert.bitpix()` for examples of types.

## Other functions for changinag data
We used 3 functions in the last section which changed around `pct`: `drop_img_dim`, `zero_trans`, and `onefile`.  

I want to explain what each of these steps are doing.  

### Dropping dimensions
When you read in some data (especially from SPM) into `R` using `readNIfTI`, the data can be read in as a 4-dimensional image, with one dimension only being one.  For example, if we look at the dimensions of `img`:
```{r too_many_dim}
dim(img)
img@dim_
pixdim(img)
```
We see that the 4th dimension is 1 and the slot `dim_` denotes it's a 4-dimensional object (4 is in the first index).  Also, we note that values in the `dim_` slot are $0$, which is not allowed when writing out:
```{r failwrite}
writeNIfTI(img, filename = tempfile())
```
In order to force it to be a 3-dimensional object, `drop_img_dim` changes all these aspects for the user.
```{r rm_dim}
img_3d = drop_img_dim(img)
dim(img_3d)
img_3d@dim_
pixdim(img_3d)
tfile = tempfile()
writeNIfTI(img_3d, filename = tfile)
```

### Unscaling the data
Many times this is not necessary, but some NIfTI files have the `scl_slope` and `scl_inter` slots defined.  These denote what to multiply and then add to the data to get it in the correct values.  `zero_trans` simply makes `scl_slope = 0` and `scl_inter = 1`.  

### Changing NIfTI magic slots
In each NIfTI image, there is a `magic` slot (see [http://nifti.nimh.nih.gov/nifti-1/documentation/nifti1fields/nifti1fields_pages/magic.html](http://nifti.nimh.nih.gov/nifti-1/documentation/nifti1fields/nifti1fields_pages/magic.html)):
```{r}
img_3d@magic
img_3d@vox_offset
```
The NIfTI documentation states:
> "ni1" means that the image data is stored in the ".img" file corresponding
   to the header file (starting at file offset 0).

We previously wrote out the 3D image, but if we try to read it back in, `readNIfTI` fails:
```{r}
readNIfTI(fname = tfile)
```
We can change the default in `readNIfTI` to `onefile = FALSE`, but nothing changes:

```{r}
writeNIfTI(img_3d, filename = tfile, onefile = FALSE)
readNIfTI(fname = tfile)
```


If we look in `oro.nifti:::.read.nifti.content` we see

>     if (!onefile) {
        if (nim@magic != "ni1") {
            stop("This is not a two-file NIfTI format")
        }

We can change this `magic` to `n+1`, but we need to change the `vox_offset = 352` and that's exactly what `onefile()` does:
```{r}
img_3d = onefile(img_3d)
writeNIfTI(img_3d, filename = tfile, onefile = FALSE)
readNIfTI(fname = tfile)
```

Viola!  All these fun little quirks that I hope you don't have to encounter.  But if you do, I hope this page helps.



