Here was a tip for authenticating a twitter account so you can still make a [TwitterMap](http://simplystatistics.org/2011/12/21/an-r-function-to-map-your-twitter-followers/)

Go to [Twitter Apps](https://dev.twitter.com/apps) and create new application - need to name it and give it a website for it, but it must be unique (to all developers, not just you).

Below, (harvested from [http://davetang.org/muse/2013/04/06/using-the-r_twitter-package/](http://davetang.org/muse/2013/04/06/using-the-r_twitter-package/)) needs to be run.  The key YOULLGETTHISFROMTWITTER when you get the app made and go under my applications.  Check out [https://dev.twitter.com/apps/new](https://dev.twitter.com/apps/new) if you can't find it.

```{r, eval=FALSE}
require("ROAuth")
require("twitteR")

#necessary step for Windows
download.file(url="http://curl.haxx.se/ca/cacert.pem", destfile="cacert.pem")

#to get your consumerKey and consumerSecret see the twitteR documentation for instructions
cred <- OAuthFactory$new(consumerKey='x8L9jCRTYiMUY6lCbQOBw',
                         consumerSecret='CZCtKDClVSRJl0LBRpABOvLZQ5MviwY6wqp1MxTMqE',
                         requestURL='https://api.twitter.com/oauth/request_token',
                         accessURL='http://api.twitter.com/oauth/access_token',
                         authURL='http://api.twitter.com/oauth/authorize')

#necessary step for Windows
cred$handshake(cainfo="cacert.pem")
```

During this spot, you need to authenticate the key by going to a website and then entering it into `R`.  Think of this as an `ask=TRUE` scenario.



```{r, eval=FALSE}
#save for later use
save(cred, file="~/twitter_authentication.Rdata")
```
Now, don't re-run the code above, just save the credentials so you don't have to do again.

```{r, eval=TRUE, echo=FALSE}
require("ROAuth")
require("twitteR")
load(file="~/twitter_authentication.Rdata")
```

```{r, eval=TRUE}
registerTwitterOAuth(cred)
```


Now, once that is done, you can go back to creating twitter maps to your heart's content:

```{r}
source("http://biostat.jhsph.edu/~jleek/code/twitterMap.R")
```

```{r, echo=FALSE}
twitterMap <- function(userName,userLocation=NULL,nMax = 1000,plotType=c("followers","both","following")){
     
  # Get location data
  cat("Getting data from Twitter, this may take a moment.\n")
  tmp = getUser(userName)
  if(is.null(userLocation)){
    userLocation = location(tmp)
    userLocation = trim(userLocation)
    if(nchar(userLocation) < 2){stop("We can not find your location from Twitter")}
  }

  followers=tmp$getFollowers(n=nMax)
  followersLocation = sapply(followers,function(x){location(x)})
  following = tmp$getFriends(n=nMax)
  followingLocation = sapply(following,function(x){location(x)})


  # Load the geographic data
  data(world.cities)
  data(us.cities)
  data(canada.cities)

  # Find the latitude and longitude of the user
  cat("Getting geographic (latitude/longitude) of Twitter users.\n")
  userLL <- findLatLon(userLocation)$latlon
  if(any(is.na(userLL))){stop("We can't find the latitude and longitude of your location from Twitter")}


  # Find the latitude and longitude of each of the followers/following
  # and calcualte the distance to the user
  
  followersLL = matrix(NA,nrow=length(followers),ncol=4)
  followingLL = matrix(NA,nrow=length(following),ncol=4)

  for(i in 1:length(followers)){
    if(length(followersLocation[[i]]) > 0){
      tmpLL = findLatLon(trim(followersLocation[[i]]))
      if(any(!is.na(tmpLL$latlon))){
        followersLL[i,] = c(unlist(tmpLL$latlon),distCosine(userLL,tmpLL$latlon),unlist(tmpLL$cont))
      }
    }
  }

   for(i in 1:length(following)){
    if(length(followingLocation[[i]]) > 0){
      tmpLL = findLatLon(trim(followingLocation[[i]]))
      if(any(!is.na(tmpLL$latlon))){
        followingLL[i,] =  c(unlist(tmpLL$latlon),distCosine(userLL,tmpLL$latlon),unlist(tmpLL$cont))
      }
    }
  }

  followingLL = followingLL[order(-followingLL[,3]),]
  followersLL = followersLL[order(-followersLL[,3]),]

  followingLL = followingLL[!is.na(followingLL[,1]),]
  followersLL = followersLL[!is.na(followersLL[,1]),]
  

  cat("Plotting results.\n")
  # Set up the colors
  cols = brewer.pal(7,"Set2")

  # Both followers and following
  if(plotType=="both"){
    data(worldMapEnv)
    par(mfrow=c(2,1),mar=rep(0,4))
    map('world',col="#191919",bg="black",fill=T,mar=rep(0,4),border=0)

    mtext(paste("@",userName," Follower Map",sep=""),col="lightgrey")
    nFollowers = dim(followersLL)[1]
    for(i in 1:nFollowers){
      greatC = getGreatCircle(userLL,followersLL[i,1:2])
      lines(greatC,col=cols[followersLL[i,4]],lwd=0.8)
    }
    
    legend(-180,0,legend = c(paste("Asia",sum(followersLL[,4]==1)),paste("Africa",sum(followersLL[,4]==2)),paste("N. America",sum(followersLL[,4]==3)),paste("S. America",sum(followersLL[,4]==4)),paste("Australia/N.Z.",sum(followersLL[,4]==5)),paste("Europe",sum(followersLL[,4]==6))),text.col=cols[1:6],bg="black",cex=0.75)
    
 
    map('world',col="#191919",bg="black",fill=T,mar=rep(0,4),border=0)
    mtext(paste("@",userName," Following Map",sep=""),col="lightgrey")
    nFollowing = dim(followingLL)[1]
    for(i in 1:nFollowing){
      greatC = getGreatCircle(userLL,followingLL[i,1:2])
      lines(greatC,col=cols[followingLL[i,4]],lwd=0.8)
    }
    
    legend(-180,0,legend = c(paste("Asia",sum(followingLL[,4]==1)),paste("Africa",sum(followingLL[,4]==2)),paste("N. America",sum(followingLL[,4]==3)),paste("S. America",sum(followingLL[,4]==4)),paste("Australia/N.Z.",sum(followingLL[,4]==5)),paste("Europe",sum(followingLL[,4]==6))),text.col=cols[1:6],bg="black",cex=0.75)

    mtext("Created by @simplystats twitterMap",side=1,adj=1,cex=0.8,col="grey")
  }

  ## Just followers
  if(plotType=="followers"){
    data(worldMapEnv)
    map('world',col="#191919",bg="black",fill=T,mar=rep(0,4),border=0)

    mtext(paste("@",userName," Follower Map",sep=""),col="lightgrey")
    nFollowers = dim(followersLL)[1]
    for(i in 1:nFollowers){
      greatC = getGreatCircle(userLL,followersLL[i,1:2])
      lines(greatC,col=cols[followersLL[i,4]],lwd=0.8)
    }
    
    legend(-180,0,legend = c(paste("Asia",sum(followersLL[,4]==1)),paste("Africa",sum(followersLL[,4]==2)),paste("N. America",sum(followersLL[,4]==3)),paste("S. America",sum(followersLL[,4]==4)),paste("Australia/N.Z.",sum(followersLL[,4]==5)),paste("Europe",sum(followersLL[,4]==6))),text.col=cols[1:6],bg="black",cex=0.75)
    mtext("Created by @simplystats twitterMap",side=1,adj=1,cex=0.8,col="grey")
 
  }

  ## Just following
  if(plotType=="following"){
    data(worldMapEnv)
    map('world',col="#191919",bg="black",fill=T,mar=rep(0,4),border=0)
    mtext(paste("@",userName," Following Map",sep=""),col="lightgrey")
    nFollowing = dim(followingLL)[1]
    for(i in 1:nFollowing){
      greatC = getGreatCircle(userLL,followingLL[i,1:2])
      lines(greatC,col=cols[followingLL[i,4]],lwd=0.8)
    }
    
    legend(-180,0,legend = c(paste("Asia",sum(followingLL[,4]==1)),paste("Africa",sum(followingLL[,4]==2)),paste("N. America",sum(followingLL[,4]==3)),paste("S. America",sum(followingLL[,4]==4)),paste("Australia/N.Z.",sum(followingLL[,4]==5)),paste("Europe",sum(followingLL[,4]==6))),text.col=cols[1:6],bg="black",cex=0.75)

    mtext("Created by @simplystats twitterMap",side=1,adj=1,cex=0.8,col="grey")
    
  }
  
}

```
and see the results!

```{r}
twitterMap("strictlystat", userLocation="Baltimore", plotType="both")
```

