```{r label=opts, results='hide', echo=FALSE, message = FALSE, warning=FALSE}
gd = getwd()
library(knitr)
opts_chunk$set(echo=TRUE, prompt=FALSE, message=TRUE, warning=TRUE, comment="", cache = TRUE)
```

The data for this post is located at [https://github.com/muschellij2/FSLR_data](https://github.com/muschellij2/FSLR_data), and was graciously allowed by our colleagues who collected it at the NIH.  

# Goal of the post
The overall goal of this post is to present a complete preprocessing pipeline for structural magnetic resonance imaging (sMRI) data "completely" within `R`.  

## Required packages
This package will rely mainly on 4 packages:

- oro.nifti: read/write data, the nifti object
- fslr: process data (need FSL for most of the functionality)
- ANTsR: process data (need cmake to install)
- extrantsr: extends and wraps ANTsR functions (need ANTsR)

## Install these packages

```{r, eval=FALSE}
install.packages("oro.nifti")
install.packages("devtools")
devtools::install_github("stnava/ANTsR") # takes a long time
devtools::install_github("muschellij2/extrantsr")
devtools::install_github("muschellij2/fslr")
```


### fslr setup
For interactive sessions, you must set the `fsl.path` to the path to FSL.  This allows fslr to execute FSL commands.  You can also set `fsl.outputtype`, which specifies the output type for images.

```{r fslr_setup, cache=FALSE}
library(fslr)
options(fsl.path="/usr/local/fsl")
options(fsl.outputtype = "NIFTI_GZ")
```


## Files

The dataset is multi-sequence MRI data where the sequences are:  

- 4 sequences: T1, T2, PD, FLAIR
- 2 time points Baseline and Follow-up

```{r, echo=FALSE}
files = c(base_t1="01-Baseline_T1.nii.gz", 
          base_t2="01-Baseline_T2.nii.gz",
          base_pd="01-Baseline_PD.nii.gz", 
          base_flair="01-Baseline_FLAIR.nii.gz", 
          f_t1= "01-Followup_T1.nii.gz", 
          f_t2= "01-Followup_T2.nii.gz", 
          f_pd = "01-Followup_PD.nii.gz", 
          f_flair="01-Followup_FLAIR.nii.gz")
```

```{r}
files
```

```{r data_setup, echo=FALSE, cache=FALSE}
fn = names(files)
homedir = path.expand("~/Dropbox/Presentations/structural_talk/")
files = file.path(homedir, files)
names(files) = fn
# setwd(homedir)
```

## Read in the Files!

Let's read in the baseline T1.  The `readNIfTI` function from the `oro.nifti` package is the workhorse:
```{r read_t1, message=FALSE}
library(oro.nifti)
base_t1 = readNIfTI(files["base_t1"])
```

## Bias Field Correction

We can do bias field correction from the `ANTsR` package.  The `bias_correct` function can perform the correction on either `nifti` objects or filenames and the user can pass in either `"N3"` or `"N4"` as an argument.  Here we will do N4 bias field correction.

```{r bias_correct, cache=TRUE}
library(ANTsR)
library(extrantsr)
n4_t1 = bias_correct(file = base_t1, correction = "N4", retimg=TRUE)
```

## Skull Stripping
We will use FSL's BET for the corrected T1 image to skull strip the image:

```{r bet_eval, cache=TRUE, echo = FALSE}
ss_t1 = fslbet(n4_t1, retimg = TRUE, opts = "-v", 
               betcmd = "bet2", 
               outfile = file.path(homedir, "SS_Image"))
fslbin(file = file.path(homedir, "SS_Image"), 
       outfile = file.path(homedir, "Brain_Mask"))
```

```{r bet, eval=FALSE}
ss_t1 = fslbet(n4_t1, retimg = TRUE, opts = "-v", 
               betcmd = "bet2", outfile = "SS_Image")
fslbin(file = "SS_Image", outfile = "Brain_Mask")
```

- `fslbet` - the command
- `retimg` - Return an image
- `opts` - options of bash `bet` command
- `betcmd` - can use `bet2` or `bet` (wrapper for `bet2`)

## Image processing

`extrantsr::preprocess_mri_within` will do N3/N4 Correction, skull strip (or mask), and register to the first scan.

```{r proc_eval, echo=FALSE, eval=TRUE, cache = TRUE}
outfiles = gsub("[.]nii", "_N4_reg.nii", files)
ofiles = outfiles[c("base_t1", "base_t2", "base_pd", "base_flair")]
if (!all(file.exists(ofiles))){
  proc_images = preprocess_mri_within(
    files = files[c("base_t1", "base_t2", "base_pd", "base_flair")],
    outfiles = ofiles,
    correct = TRUE,
    correction = "N4",
    retimg = TRUE, maskfile = file.path(homedir, "Brain_Mask.nii.gz"))
}
```

```{r proc, eval=FALSE}
outfiles = gsub("[.]nii", "_N4_reg.nii", files)
proc_images = preprocess_mri_within(
  files = files[c("base_t1", "base_t2", "base_pd", "base_flair")],
  outfiles = outfiles[c("base_t1", "base_t2", "base_pd", "base_flair")],
  correct = TRUE,
  correction = "N4",
  retimg = TRUE, maskfile = "Brain_Mask.nii.gz")
```

## Intensity Normalization

```{r whitestripe_eval, eval=TRUE, echo=FALSE, cache=TRUE}
wsfiles = gsub("[.]nii", "_N4_reg_whitestripe.nii", files)
wsfiles = wsfiles[c("base_t1", "base_t2", "base_pd", "base_flair")]
ws.outfile = file.path(homedir, "WS_Mask.nii.gz")
if (!all(file.exists(c(wsfiles, ws.outfile)))){
  results = reg_whitestripe(t1 = outfiles["base_t1"], 
                  register = TRUE, 
                  native = TRUE,
                  t1.outfile = wsfiles["base_t1"],
                  other.files = outfiles[c("base_t2", 
                                           "base_pd", "base_flair")],
                  other.outfiles = wsfiles[c("base_t2", 
                                             "base_pd", "base_flair")],
                  ws.outfile = ws.outfile,
                  mask = file.path(homedir, "Brain_Mask.nii.gz"))
}
```

```{r whitestripe, eval=FALSE}
wsfiles = gsub("[.]nii", "_N4_reg_whitestripe.nii", files)
results = reg_whitestripe(t1 = outfiles["base_t1"], 
                register = TRUE, 
                native = TRUE,
                t1.outfile = wsfiles["base_t1"],
                other.files = outfiles[c("base_t2", 
                                         "base_pd", "base_flair")],
                other.outfiles = wsfiles[c("base_t2", 
                                            "base_pd", "base_flair")],
                ws.outfile = "WS_Mask.nii.gz",
                mask = "Brain_Mask.nii.gz")
```




## Skull Stripping
We will use FSL's BET for the corrected T1 image to skull strip the image:

```{r f_bet_eval, cache=TRUE, echo = FALSE}
f_t1 = readNIfTI(files["f_t1"])
n4_t1 = bias_correct(file = f_t1, correction = "N4", retimg=TRUE)
f_ss_t1 = fslbet(n4_t1, retimg = TRUE, opts = "-v", 
               betcmd = "bet2", outfile = file.path(homedir, "F_SS_Image"))
fslbin(file = file.path(homedir, "F_SS_Image"), outfile = file.path(homedir, "F_Brain_Mask"))
```

## Image processing

`extrantsr::preprocess_mri_within` will do N3/N4 Correction, skull strip (or mask), and register to the first scan.

```{r f_proc_eval, echo=FALSE, eval=TRUE, cache = TRUE}
outfiles = gsub("[.]nii", "_N4_reg.nii", files)
infiles = files[c("f_t1", "f_t2", "f_pd", "f_flair")]
ofiles = outfiles[c("f_t1", "f_t2", "f_pd", "f_flair")]
if (!all(file.exists(ofiles))){
  proc_images = preprocess_mri_within(
    files = infiles,
    outfiles = ofiles,
    correct = TRUE,
    correction = "N4",
    retimg = TRUE, maskfile = file.path(homedir, "F_Brain_Mask.nii.gz"))
}
```


## Intensity Normalization

```{r f_whitestripe_eval, eval=TRUE, echo=FALSE, cache=TRUE}
wsfiles = gsub("[.]nii", "_N4_reg_whitestripe.nii", files)
f_wsfiles = wsfiles[c("f_t1", "f_t2", "f_pd", "f_flair")]
ws.outfile = file.path(homedir, "F_WS_Mask.nii.gz")
if (!all(file.exists(c(wsfiles, ws.outfile)))){
  results = reg_whitestripe(t1 = outfiles["f_t1"], 
                            register = TRUE, 
                            native = TRUE,
                            t1.outfile = f_wsfiles["f_t1"],
                            other.files = outfiles[c("f_t2", 
                                                     "f_pd", "f_flair")],
                            other.outfiles = wsfiles[c("f_t2", 
                                                       "f_pd", "f_flair")],
                            ws.outfile = ws.outfile,
                            mask = file.path(homedir, "F_Brain_Mask.nii.gz"))
  }
```


```{r coreg, eval=TRUE, echo=FALSE, cache=TRUE}
outfiles = gsub("[.]nii", "_N4_reg.nii", files)
wsfiles = gsub("[.]nii", "_N4_reg_whitestripe.nii", files)
f_ws.mask = file.path(homedir, "F_WS_Mask.nii.gz")
f_ws.maskout = file.path(homedir, "F_WS_Mask_tobase.nii.gz")
f_wsfiles = wsfiles[c("f_t1", "f_t2", "f_pd", "f_flair")]
f_wsfiles.out = gsub("[.]nii", "_tobase.nii", f_wsfiles)
f_outfiles = outfiles[c("f_t1", "f_t2", "f_pd", "f_flair")]
f_outfiles.out = gsub("[.]nii", "_tobase.nii", f_outfiles)
if (!all(file.exists(c(f_wsfiles.out, f_outfiles.out, f_ws.maskout)))){
  res = ants_regwrite(filename = outfiles["f_t1"], 
              outfile = NULL,
              template.file = outfiles["base_t1"], 
              typeofTransform = "Rigid",
              interpolator = "LanczosWindowedSinc",
              other.files = c(f_wsfiles, 
                              f_outfiles, f_ws.mask),
              other.outfiles = c(f_wsfiles.out, 
                                 f_outfiles.out, f_ws.maskout), 
              skull_strip = FALSE, 
              correct = FALSE, 
              retimg = TRUE,
              remove.warp = TRUE)
}
fslthresh(file = f_ws.maskout, outfile = f_ws.maskout, 
          thresh = 0.5)
fslbin(file = f_ws.maskout, outfile = f_ws.maskout)
```

```{r joint_norm, eval=TRUE, echo=FALSE, cache=TRUE}
library(WhiteStripe)
joint = file.path(homedir, "Joint_WS_Mask.nii.gz")
ws.base = file.path(homedir, "WS_Mask.nii.gz")
ws.fup = file.path(homedir, "F_WS_Mask_tobase.nii.gz")
joint = fslmask(file = ws.base, 
                mask = ws.fup, 
                outfile = joint, retimg = TRUE)
indices = which(joint > 0)

wsfiles = gsub("[.]nii", "_N4_reg_whitestripe.nii", files)
wsfiles[c("f_t1", "f_t2", "f_pd", "f_flair")] = gsub("[.]nii", "_tobase.nii", wsfiles[c("f_t1", "f_t2", "f_pd", "f_flair")])
joint.ws = gsub("[.]nii", "_joint.nii", wsfiles)
runfiles = outfiles
runfiles[c("f_t1", "f_t2", "f_pd", "f_flair")] = f_outfiles.out
brainmask = readNIfTI(file.path(homedir, "Brain_Mask.nii.gz"),
                      reorient=FALSE)
dtype = function(img){
  img = drop_img_dim(img)
  img = datatyper(img, 
                 datatype= convert.datatype()$FLOAT32,
                 bitpix= convert.bitpix()$FLOAT32)
  img = cal_img(img)
  return(img)
}
ifile = 1
for (ifile in seq_along(runfiles)){
  if (!file.exists(joint.ws[ifile])){
    img = readNIfTI(runfiles[ifile], reorient = FALSE)
    img = whitestripe_norm(img, indices)
    img = mask_img(img, brainmask)
    img = dtype(img)
    writeNIfTI(nim = img, filename = nii.stub(joint.ws[ifile]))
  }
  print(ifile)
}
```


```{r diffs, eval=TRUE, echo=FALSE, cache=TRUE}
df = data.frame(type = c("t1", "t2", "pd", "flair"), stringsAsFactors = FALSE)
df$base = joint.ws[paste0("base_", df$type)]
df$f = joint.ws[paste0("f_", df$type)]
rownames(df) = df$type
df$type = NULL
brainmask = readNIfTI(file.path(homedir, "Brain_Mask.nii.gz"),
                      reorient=FALSE)
brainmask[brainmask == 0] = NA
itype = "t2"
base = readNIfTI(df[itype, "base"], reorient = FALSE)
f = readNIfTI(df[itype, "f"], reorient = FALSE)
ddf = data.frame(base = c(base), fup = c(f))
ddf = ddf[ brainmask %in% 1, ]
samp = ddf[ sample(nrow(ddf), size=1e5), ]
d = niftiarr(base, base - f)
d = mask_img(d, mask = brainmask)
```

```{r tryout, eval = TRUE}
outfiles = gsub("[.]nii", "_N4_reg.nii", files)
f_outfiles = outfiles[c("f_t1", "f_t2", "f_pd", "f_flair")]
f_outfiles = gsub("[.]nii", "_tobase.nii", f_outfiles)
b_outfiles = outfiles[c("base_t1", "base_t2", "base_pd", "base_flair")]
df = data.frame(type = c("t1", "t2", "pd", "flair"), stringsAsFactors = FALSE)
df$base = b_outfiles[paste0("base_", df$type)]
df$f = f_outfiles[paste0("f_", df$type)]
rownames(df) = df$type
df$type = NULL
brainmask = readNIfTI(file.path(homedir, "Brain_Mask.nii.gz"),
                      reorient=FALSE)
itype = "t2"
base = readNIfTI(df[itype, "base"], reorient = FALSE)
base = zscore_img(base, mask = brainmask, margin = NULL)
f = readNIfTI(df[itype, "f"], reorient = FALSE)
f = zscore_img(f, mask = brainmask, margin = NULL)
ddf = data.frame(base = c(base), fup = c(f))
ddf = ddf[ brainmask %in% 1, ]
samp = ddf[ sample(nrow(ddf), size=1e5), ]
d = niftiarr(base, base - f)
d = mask_img(d, mask = brainmask)
ortho2(d)
```


## Other useful functions from fslr

- `window_img` - set min and max values of image (good for plotting)
- `drop_img_dim` - drop dimension if is dimension is 1, especially if you have the error `Error: all dim elements > dim[1] must be 1 dim/pixdim mismatch`
- `mask_img` - mask image with an array/nifti object
- `robust_window` - quantile the image then use `window_img` (take out large values)
- `fslhd` and `fslval` can get information from an image without having to read it in
- `checkimg` - checks if an image is character, writes `nifti` to tempfile, returns character
- `check_nifti` - checks if an image is `nifti`, reads filename, returns `nifti`

## Overview

1. `oro.nifti` - reading/writing nifti objects 
2. `ANTsR`
    + bias field corrrection
    + registration: linear, non-linear (symmetric)
3. `fslr`
    + can perform image operations from FSL
    + has helper functions for `nifti` objects
4. extrantsr - helper functions for ANTsR
    + wrapper functions for preprocessing

```{r, echo=FALSE}
setwd(gd)
```



```{r whitestripe_t2_eval, eval=TRUE, echo=FALSE, cache=TRUE}
t2wsfiles = gsub("[.]nii", "_N4_reg_whitestripe_t2.nii", files)
t2wsfiles = t2wsfiles[c("base_t1", "base_t2", "base_pd", "base_flair")]
t2ws.outfile = file.path(homedir, "T2_WS_Mask.nii.gz")
if (!all(file.exists(c(t2wsfiles, t2ws.outfile)))){
  results = reg_whitestripe(
    t1 = outfiles["base_t1"],
                            type = "T2",
                            t1.outfile = tempfile(fileext = ".nii.gz"),
                            t2 = outfiles["base_t2"], 
                            register = TRUE, 
                            native = TRUE,
                            t2.outfile = t2wsfiles["base_t2"],
                            other.files = outfiles[c("base_t1", 
                                                     "base_pd", "base_flair")],
                            other.outfiles = t2wsfiles[c("base_t1", 
                                                         "base_pd", "base_flair")],
                            ws.outfile = t2ws.outfile,
                            mask = file.path(homedir, "Brain_Mask.nii.gz"))
}
```