# Followup on hold jobs
<link rel="stylesheet" href="http://yandex.st/highlightjs/7.3/styles/default.min.css">
<script src="http://ajax.googleapis.com/ajax/libs/jquery/1.9.0/jquery.min.js"></script>
<script src="http://yandex.st/highlightjs/7.3/highlight.min.js"></script>

In this post I have found a way to hold a pipeline from excecuting if any of the preceding jobs have errored.  
I created a quick `R` file, named it `test.R`.  Contents of `test.R`:
```{r, comment="", error=TRUE, highlight=TRUE}
print(x)
```
Most of you can see that this is destined to fail since `x` can't be found, which is the point.  I want it to fail.  Now I created my bash file, named it `test.sh`, with contents below:
```{r, engine='bash', eval=FALSE, highlight=TRUE}
#!/bin/bash 
R --no-save < test.R
if [ "$?" -gt "0" ]
then
  exit 100
fi
```
This program simply says run this `R` file, then check the last exit status in `bash` saved as `$?`, see if its greater than `0`.  If this is greater than `0`, then `R` file failed.  Then just exit `bash` with exit code `100`.  Now, why use exit code `100`? If you do 
```{r, engine='bash', eval=FALSE}
man qsub
```
you'll notice that under `-hold_jid` they have noted:

>If any array task of the referenced jobs exits with exit code 100, the dependent tasks of the submitted job will remain  ineligible  for  execution.

so we need to tell the Sun Grid Engine (SGE) that there is an exit code of `100` if we want it to hold up our jobs.  

### What happens with held jobs?
What happens when we try something like
```{r, engine='bash', eval=FALSE, comment="", echo=TRUE}
#!/bin/bash 
qsub -cwd -N test1 test.sh 
qsub -cwd -hold_jid test1 -N test2 test.sh
```
and then things fail?  We get the following output:
```
job-ID  prior   name       user         state submit/start at     queue                          slots ja-task-ID 
-----------------------------------------------------------------------------------------------------------------
 990025 0.58333 test1      XXXXXXXX     Eqw   11/19/2013 16:54:56                                    1        
 990026 0.00000 test2      XXXXXXXX     hqw   11/19/2013 16:54:56                                    1
```
the `E` denotes tthere was an error.  The `test2` job will not run and will stay held (unless you delete the `test1` job using `qdel`).  

Now, the <strong>big</strong> problem with this is that it tells our cluster administrator that your job has failed, which is not really the best option.  We may work out a system where we can put a special naming convention so that the failure can be filtered out compared to true failures (such as node crashes).  

### Alternative option not using exit code `100`

But for now, you could do something simple like at the beginning of pipline/series of `sh` scripts:
```{r, engine='bash', eval=FALSE}
echo 0 > check.txt
```
then in each `sh` script you can do:
```{r, engine='bash', eval=FALSE}
while read line           
do           
  if [ "$line" -gt "0" ]
  then
	  exit 1
	fi
done < check.txt 
R --no-save < test.R
echo $? >> check.txt
```
so that any time you get an exit status > 0, then you can stop everything down stream.  This works and is considered somewhat of a "blackboard" approach in that jobs check a blackboard before continue running.  

Potential problems: 
  * Needing to cleanup the check.txt file
  * Multiple jobs reading the file at the same time (race condition).  This may be mitigated if you have a pause command after the last `echo`.  
  * You don't have a record of what file actually failed (or which array index) - you can `cat`/`echo` more to the file and then use `awk` or something else to just take the first column (exit code).
    
But for now, I'll just try to make sure things run smoothly interactively for a bit, maybe doing some of [this](http://xkcd.com/303/).
