```{r label=opts, results='hide', echo=FALSE, message = FALSE, warning=FALSE}
library(knitr)
opts_chunk$set(echo=TRUE, prompt=FALSE, message=FALSE, warning=FALSE, comment="", results='hide')
```

Recently, [Jeff Leek](http://jtleek.com/) had proposed something at tea time:

> Why should I switch over to [ggplot2](http://ggplot2.org/)? I can do everything in base graphics.

I have heard this argument before and I understand it for the most part.  Many people have learned `R` before `ggplot2` came on the scene (as well as [many other packages](http://kbroman.org/hipsteR/)) and learned to do all the things they needed to.  Some don't understand the new syntax for plotting and argue the learning curve is not worth the effort.  Also, many say it's not straightforward to customize.


As the discussion progressed, I told him to send me 10 plots that he commonly makes, and I would recreate them in `ggplot2`.  The goal was to help him break into `ggplot2` if he wanted, and also see if they were "as easy" as doing them in base.  I want to discuss my results and the fact that `ggplot2` is not ALWAYS the answer, nor was it supposed to be. 

## First Plot - Heatmap

The first example Jeff wanted a heatmap that had 100,000 rows and 100 columns with a good default color scheme.  As a comparison, I used `heatmap` in base `R`.  Let's do only 10,000 to start:

```{r heatmap, eval = TRUE, cache=TRUE, results='markup'}
N = 1e4
mat = matrix(rnorm(100*N), nrow=N, ncol=100)
colnames(mat) = paste0("Col", seq(ncol(mat)))
rownames(mat) = paste0("Row", seq(nrow(mat)))
system.time({heatmap(mat)})
```


For a heatmap in `ggplot2`, I used `geom_tile` and wanted to look at the results.  Note, `ggplot2` requires the data to be in "long" format, so I had to reshape the data.

```{r gtile, eval = TRUE, cache=TRUE, results='markup'}
library(reshape2)
library(ggplot2)
df = melt(mat, varnames = c("row", "col"))
system.time({
  print({ 
    g= ggplot(df, aes(x = col, y = row, fill = value)) + 
      geom_tile()}) 
  })
```

One of the problems is that `heatmap` does clustering by default, so the comparison is not really fair (but still using "defaults" - as Jeff specified).  Let's do the `heatmap` again without the clustering.

```{r heatmap_alone, eval = TRUE, cache=TRUE, results='markup'}
system.time({heatmap(mat, Rowv = NA, Colv = NA)})
```

Which one looks better? I'm not really sure, but **I** do like the red/orange coloring scheme - I can differences a bit better.  Granted, there shouldn't be many differences as the data is random.  The `ggplot` graph does have a built-in legend of the values, which is many times necessary.  Note, the rows of the `heatmap` are shown as columns and the columns shown as rows, a usually well-known fact for users of the `image` function.  The `ggplot` graph plots as you would a scatterplot -- up on x-axis and right on the y-axis are increasing.  If we want to represent rows as rows and columns as columns, we can switch the `x` and `y` aesthetics, but I wanted as close to `heatmap` as possible.

### Don't factor the rows
Note, I named the matrix rows and columns, if I don't do this, the plotting will be faster, albeit slightly.

```{r heatmap2_2, eval = TRUE, cache=TRUE, results='markup'}
N = 1e4
mat = matrix(rnorm(100*N), nrow=N, ncol=100)
df = melt(mat, varnames = c("row", "col"))

system.time({heatmap(mat, Rowv = NA, Colv = NA)})
system.time({
  print({ 
    g= ggplot(df, aes(x = col, y = row, fill = value)) + 
      geom_tile()}) 
  })
```

### 20,000 Observations
I'm going to double it and do 20,000 observations and again not do any clustering:

```{r heatmap2, eval = TRUE, cache=TRUE, results='markup'}
N = 2e4
mat = matrix(rnorm(100*N), nrow=N, ncol=100)
colnames(mat) = paste0("Col", seq(ncol(mat)))
rownames(mat) = paste0("Row", seq(nrow(mat)))
df = melt(mat, varnames = c("row", "col"))

system.time({heatmap(mat, Rowv = NA, Colv = NA)})
system.time({
  print({ 
    g= ggplot(df, aes(x = col, y = row, fill = value)) + 
      geom_tile()}) 
  })
```

### 100,000 Observations
Let's scale up to the 100,000 observations Jeff requested.

```{r heatmap3, eval = TRUE, cache=TRUE, results='markup'}
N = 1e5
mat = matrix(rnorm(100*N), nrow=N, ncol=100)
colnames(mat) = paste0("Col", seq(ncol(mat)))
rownames(mat) = paste0("Row", seq(nrow(mat)))
df = melt(mat, varnames = c("row", "col"))

system.time({heatmap(mat, Rowv = NA, Colv = NA)})
system.time({
  print({ 
    g= ggplot(df, aes(x = col, y = row, fill = value)) + 
      geom_tile()}) 
  })
```


We see that `heatmap` and `geom_tile()` scale with the number observations on how long it takes to plot, but `heatmap` being **much** quicker.  There may be better ways to do this in `ggplot2`, but after looking around, it looks like `geom_tile()` is the main recommendation.  Overall, for doing heatmaps of this size, I would use the `heatmap`, after transposing the matrix and use something like `seq_gradient_pal` from the `scales` package or `RColorBrewer` for color mapping.  

For smaller dimensions, I'd definitely use `geom_tile()`, especially if I wanted to do something like [map text to the blocks as well](https://www.biostars.org/p/56291/).  The other benefit is that no transposition needs to be done, but the data does need to be reshaped explicitly.  


Like I said, `ggplot2` is not ALWAYS the answer; nor was it supposed to be.  


## ggplot2 does not make publication-ready figures
Another reason for the lengthy discussion is that Jeff states **ggplot2 does not make publication-ready figures**.  I agree.  Many new users need a large
> ggplot2 does not make publication-ready figures

message with their first install.  But **neither does base graphics**.  Maybe they need that message for their first `R` install.  R would boot up, and you can't start until you answer "Does R (or any statistical software) give publication-ready figures by default".  Maybe, if you answer yes, R self-destructs.

Overall, a publication-ready figure takes time, customization, consideration about point size, color, line types, and other aspects of the plot, and usually stitching together multiple plots.  `ggplot2` has a lot of default features that have given considerate thought to color and these other factors, but one size does not fit all.  The default color scheme may not be consistent with your other plots or the data.  If it doesn't work, change it.

And in this I agree with Jeff: many people make a plot in `ggplot2` and it looks good and they do not put the work in to make it publication-ready.  How many times have you seen the muted green and red/pink color that is the default first 2 colors in a `ggplot2` plot?


## Why all the ggplot2 hate?
I'm not hating on `ggplot2`; I use it often and value it.  My next post will be about plots I commonly use in `ggplot2` and why I chose to use it. I think it makes graphs that by default look better than many base alternatives.  It does things many things well and easier than base.  It has good defaults for some things.  It's a different grammar that, when learned, is easier for plotting.  However, I do not think that `ggplot2` figures are ready-set-go for publication.

Although I think `ggplot2` is great, I most-definitely think base `R` graphics are useful. 

`ggplot2` is not ALWAYS the answer.  Again, it was never supposed to be.

